Prevailing scenarios of functional change in Anthropocene bird and mammal communities

### 2024-03-11 paper/GEB Submission/manuscript_GEB.tex
Just... so many changes.

The meaning of general trends derived from limited data, and their
relevance for conservation, is a topic of on going debate. Global
analyses have been heavily criticized for geographic biases, lack of
data in the most heavily impacted areas, and not considering the
ecological context of individual studies (10--12). Many of these
criticisms reflect limitations of ecological data on the whole, and
proponents argue we must make the best of the imperfect data available
(13). Still, there is general consensus on a few local-scale patterns
that seem to be characteristic of the anthropocene, including the
prevalence of species turnover across communities and patterns of both
richness increase and decrease in different taxa and biomes (14).
Understanding the implications of change in species assemblage for
ecosystem stability and function is a critical next step in moving
forward the biodiversity change conversation.

Functional diversity offers a potentially powerful alternative to
species-based approaches for detecting and describing community change
by providing a mechanistic link between species' response to
environmental change (\emph{response traits}) and the processes they
perform (\emph{effect traits}) (15--17). By describing the functional
trait space, functional diversity metrics capture the disproportionate
impact of losses or gains of functionally unique species. Functional
diversity metrics are therefore particularly well suited for assessing
community shifts underlying even constant species richness trends.

Beyond simply characterizing changes in community structure, trends in
functional composition also have important implications for ecosystem
stability, function, and resilience. There is increasing evidence
functional diversity is a better predictor of ecosystem function than
species-based metrics (18, 19), and that different facets of functional
diversity play essential roles in maintaining ecosystem stability (20,
21). Almost all hypothesized mechanisms underpinning the relationship
between species diversity and ecosystem function are trait-dependent
(22). Determining functional trends therefore gives a more direct
picture of potential trends in critical ecosystem processes.

It is critical to establish whether or not functional loss is prevalent
across communities. While functional loss is frequently cited as one of
the most pressing concerns of the anthropocene (23--25), functional
diversity may be maintained even when species are lost from a community
(26, 27). Forecasts of functional loss range from negligible (28) to
dire (29, 30). And while some observed trends show significant
functional loss (31) others document no loss even in some of the most
heavily impacted communities (32--34). On paleoecological time scales
functional composition shows mixed responses to environmental change and
extinction events (35, 36), with significant impacts of species
extinctions on functional diversity in some taxa and not others (37).
For some time periods, functional structure appears to be maintained for
substantial portions of geological time (38). Still, some losses of
functional diversity are indisputable on both paleoecological and
contemporary timescales such as continued trophic downgrading due to
loss of large-bodied mammals, but implications of those losses for local
diversity patterns are less clear (39, 40) .

Assessments of broad-scale temporal change in functional diversity have
previously been limited by a lack of functional trait data. The majority
of work has therefore focused largely on system-specific studies with
traits collected \emph{in situ}. This has challenged our ability to
establish general rules for how functional composition may be changing
through time, particularly in response to global change drivers. Ongoing
efforts to assemble functional traits for a variety of taxa have made
synthesis of existing community assemblage data and functional traits
possible for the first time, providing initial insights into the ways
functional diversity changes on a broad scale for specific taxa
(e.g.~fish, 41, birds, 42, 43). However, to date there has been no
cross-taxa assessment of temporal functional change for a broad
geographic and taxonomic extent (44).

Here we perform the first multi-taxa, multi-realm assessment of
functional diversity change through time. We focus on mammal and bird
species as a significant subset of the world's biodiversity heavily
impacted by anthropogenic change. While examining trends in plants,
invertebrates, and other vertebrate species is of equal interest, trait
data for those taxa raise additional challenges such as limited and
biased species coverage (45), a lack of accepted species-level means,
and differences in the types of traits collected. To ensure
comparability across taxa in trait type and data quality we therefore
focus on mammals and birds. Traits were intentionally selected to be
representative of a species' Eltonian niche, thereby summarizing the
functional role they play in the community (46). An initial assessment
of amphibian trends is included in the supplement, but excluded from
general trend assessment here due to limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series
to determine whether or not there is a general trend of functional
change, both in observed metrics and in metrics corrected for changes in
species richness. We further assess the prevailing modes of change
occurring at the study level to determine how functional change is
related to species patterns, and which kinds of species and functional
change are most common across communities. We expect to find significant
functional changes happening alongside and independent of species
richness changes as a result of ongoing anthropogenic impacts and global
change drivers.

@@

We obtained mammal and bird time series from the BioTIME database, a
global repository of high-quality assemblage time series. All studies
included in the database follow consistent sampling protocols and
represent full assemblages rather than populations of single species
(36). All time series include abundance of observed species. Following
best practices for the database (47), studies with multiple sample
locations were split into individual time series following a
standardized spatial scale. Scale was set by a global grid with cell
size determined based on the sample extent of studies with only a single
location (see 36 for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations. All samples from a study within a single
cell were considered to be a single time series, and species abundances
were combined for all samples.

We used trait data from the Elton Trait Database, which consists of
species-level means for traits that represent species' multifaceted role
in the community (46). Traits include: body mass, diet, active diel
period, nocturnality, forest foraging strata, pelagic use. Multiple
traits (i.e.~diet, foraging strata, activity seasonality, active diel
period) were broken down into percentage or binary use for each level.

In order to ensure taxonomic consistency across datasets, BioTIME
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (retrieved
09-15-2020 from the on-line database,
\url{https://doi.org/10.5066/F7KH0KBK}), obtained through the
\texttt{taxadb} R package (48, 49). If more than one species in the
assemblage data resolved to the same identifier, observations were
considered the same species. For trait data, traits for all species of
the same identifier were averaged. Only studies for which at least 75\%
of species had trait data were included. In order to have a sufficient
number of species to calculate functional diversity metrics, years with
fewer than 5 species observed were also excluded. Sensitivity analyses
were conducted for the trait coverage threshold and the duration of
included time series.

@@

Functional diversity metrics were calculated using the \emph{dbFD}
function from the \emph{FD} R package (50). Here we report functional
richness (\emph{FRic}), functional evenness (\emph{FEve}), and
functional divergence (\emph{FDiv}) which together describe three
complementary characteristics of the functional space (22, 51).
\emph{FRic} assesses the volume of the trait space occupied by species
in the community, with higher values indicating communities with species
of more extreme trait values. \emph{FEve} describes how species are
distributed across the trait space and how abundance is distributed
across species. Higher values of \emph{FEve} indicate more even spacing
of species in the trait space and individuals across species.
\emph{FDiv} measures the degree to which species and their abundances
maximize differences in the functional space. Higher values of
\emph{FDiv} therefore correspond to communities where many highly
abundant species are on the edges of the trait space. We also calculated
the community-weighted mean (\emph{CWM}) of included traits to examine
shifts in the distribution of each trait.

All available trait data for each study were included in functional
diversity calculations with the exception of traits that were the same
value for all observed species in the study. For variables with multiple
levels each level was included as a separate trait axis. Continuous
traits were z-score scaled to give each trait equal weight in the trait
space (52,
53).\textcolor{blue}{Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA. We limited the number of included PCoA axes to}
the maximum number of traits that fulfills the criteria \(s >= 2^t\),
where \(s\) is the number of species and \emph{t} is the number of
traits. This restriction allows for enough axes to capture the trait
space while maintaining computational feasibility (54). Metrics
incorporated weighting based on species abundance.

@@

To assess functional change independent of species richness we
calculated the standardized effect size (SES) for each of the three
summary functional diversity metrics (\emph{FRic}, \emph{FEve},
\emph{FDiv}) from null estimates (55). Null model corrections allow us
to assess the degree to which the observed functional diversity metric
deviates from the value expected by chance in a randomly assembled
community. Null estimates were calculated for each rarefied sample by
randomly sampling species from the species pool for each year and
randomly assigning observed abundances to species. Species pools
included all species observed for a time series. This process was
repeated 500 times to get an estimate and standard deviation of the null
expectation for the metric for each rarefaction sample for that time
series. We used these values to calculate SES using the following
formula: \(SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}\). We
then calculated the median SES estimate for each metric from all the
rarefaction samples for a time series. SES estimates can be interpreted
as how much of the functional characteristic (richness, evenness,
divergence) was observed beyond what was expected by chance for a
community of that species richness.

@@

We estimated general trends for each diversity metric using a linear
mixed effects model with a random slope and intercept for each study and
each time series nested within the study. We fit 18 individual
\emph{CWM} models. All time series with data for a given trait were
included in the corresponding \emph{CWM} model. We estimated study level
trends using individual linear models. For studies with more than one
times series we fit a random slope and intercept for time series. Some
study-level models could not be fit for five studies for at least one
metric due to data limitations, but those studies were still included in
the general models. They represented 12 of 1350 study-level models fit
for each metric. For further details see the supplement. Where
appropriate, response variables were \(log\) or \(log(x+1)\) transformed
to better fit model assumptions of residual normality.

To test for trends within and between different levels of taxa, biome,
and realm we fit separate models with each of those factors added as a
predictor to the original model structure. We estimated within-level
slopes and calculated between-level contrasts using the \emph{emmeans}
package (v1.8.2, 56). We assessed the impact of time series duration and
start year on study-level trends using linear models with duration and
start year as predictors. All models in our analysis were fit using the
\emph{lme4} (v1.1-30) package in R (v4.2.3) and p-values were calculated
by Satterthwaite's degrees of freedom method using the \emph{lmerTest}
(v3.1-3) package with a significance level of \(\alpha = 0.05\)
(57--59).

@@

We found no significant overall trend in species richness or summary
functional diversity metrics (observed or standardized) (Fig
\ref{fig:timeseriesPlot}). We did find a significant overall decrease in
Jaccard similarity, indicating accumulating changes in species
composition. Non-significant overall trends indicate that although some
studies experience increasing or decreasing trends, the average trend
across studies was plausibly 0 (Table \ref{tab:resultsTab}). Trends for
different taxa, biomes, or realms were also non-significant for richness
and summary functional diversity metrics, with the exception of a
significantly increasing trend for functional evenness of global studies
(characterized by having samples on multiple continents), and a
significantly decreasing functional richness slope for
temperate/tropical studies and mammal studies. However, trends were not
exhibited for the standardized metric indicating that differences were
largely due to changes in species richness. Further, with only two
global studies, the trend should not be considered truly general. The
general trends for \emph{CWM} models were similarly not significant,
with a significant positive trend for only percentage of fish in diet
composition.

@@

At the study level, 15 of 50 studies exhibited a significant trend in
species richness and 11 exhibited significant turnover. For observed
functional metrics, 25 of 50 studies exhibited a trend in a least one
metric, and 15 of 50 studies exhibited a significant trend for at least
one standardized metrics (Table \ref{tab:trendTab}). In general, there
were more significant trends for uncorrected metrics, with some
disappearing after correction, indicating that those trends were likely
due to changes in the number of species. Hypothesis testing for
study-level trends is likely affected by multiple testing issues and
some trends identified as significant are therefore potentially
spurious. Rather than interpreting changes in specific studies, we
present these results as a general picture of the kinds of trends
experienced by communities.

Study-level slopes for summary functional metrics were significantly
related to start year of the time series for Jaccard similarity and
functional evenness, both of which had significantly more negative
slopes with more recent start year. No summary functional metrics were
significantly related to the start year of the time series.

@@

Our study represents the largest broad-scale multi-taxa assessment of
functional change through time to date, giving a first look at aggregate
and local trends in functional diversity in mammal and bird communities.
Surprisingly, we did not detect an overall trend in any of the summary
functional diversity metrics. As with previous species-based syntheses,
we also found no overall trend in species richness accompanied by
increasing turnover through time (36), indicating that non-significant
trends in functional metrics may be consistent with similar
well-documented species derived trends. We found no trend in functional
change for almost all realms, biomes, and taxonomic groups, with
evidence of functional richness loss in mammal studies only. These
results are consistent with multiple studies linking anthropogenic
drivers to loss of functional diversity in mammal communities (60--62)

Despite a lack of general trends in summary functional metrics, we did
find multiple trends in \emph{CWM}'s for studies coming from the same
taxa, realms, or climates, many of which are consistent with previous
findings for those contexts. For example, the decrease in insect
consumption for tropical communities reflects well documented declines
in tropical insectivorous birds (63). Reduction in utilization of
understory foraging for terrestrial communities could be the result of
disturbance like human recreation or increased predation from introduced
species. Other shifts in diet for both Birds and Mammals, including
increasing fruit, seed, and vertebrate consumption and decreasing seed
consumption in some climates point to important areas for further
exploration.

@@

About 15 percent of studies, which were also exclusively the longest
running studies, had a trend in turnover, richness, and functional
change. The studies in this scenario fell into two groups: a positive
trend in richness with a positive trend in functional richness
(corrected and uncorrected), and a negative trend in species richness
with a negative trend in uncorrected functional richness and no trend in
corrected richness but a trend in other corrected functional metrics.
This would indicate that when species are lost from a community, they
are not more functionally unique than expected by chance, though they
may produce shifts in other aspects of functional structure. When
species are gained however, they disproportionately add to the
functional richness of the community.

Our results illustrate the range of relationships between species-based
and functional metrics observed in real communities. Critically,
particularly for the short observation time windows that characterize
the majority of ecological data, functional shifts may be underlying
seemingly static species-based community metrics. Species-based
approaches may therefore be missing areas where significant change in
the functional space could be impacting ecosystem function and
resilience. In scenarios where there are significant species gains or
losses over time, we found evidence that species losses may be
relatively functionally indistinct, while species gains contribute
disproportionately to the richness of a functional space. This finding
is consistent with evidence that at least in bird communities, common
functionally general species are being lost even as rare species are
increasing (64--66).

Our results are consistent with ecological expectations that community
dissimilarity would be greater the larger the observation time window
purely due to background processes, which is likely further accelerated
by global change drivers (14). While we do not address how observed
rates differ from background expectation here, we can make some
assertions about the implication of turnover for functional change.
Where turnover is occurring, it is almost always accompanied by a
directional change in the functional space. Species are not therefore
being replaced by exact functional counter parts. The fact that we
observed both increases and decreases in corrected functional richness
accompanying turnover is a reminder that replacements are not always an
indication of functional loss but can also lead to increased functional
richness. While shifts in functional space may be the result of
environmental pressure, we cannot distinguish between directional change
and shifts due to the inavailability of an analog species in the species
pool.

@@

Still, some results stand in contrast to predictions for trait shifts
under global change. For example, mean body size is predicted to
decrease as a result of climate change impacts and megafaunal loss (67),
a phenomenon which has already been well documented empirically and
experimentally in multiple taxa (68--71). While the species-level trait
means used here are not appropriate for assessing intraspecific body
size shifts, we would expect to see shifts in community-wide means due
to local losses of large-bodied species. Instead, we found no evidence
of a trend in \emph{CWM} body size, with the exception of a single study
of marine mammals in the Bahamas which showed a significant increase.
Still, many of the studies in our dataset draw from areas that may have
experienced significant loss of large-bodied species before the
observation window, with contemporary loss rates slowing (72). Trends
could be significantly different for the same time periods in regions of
sub-Saharan Africa, for example, which has poor representation in our
dataset but where megafauna exist on the landscape and are increasingly
threatened (73).

What does local maintenance of functional structure mean for ecosystem
function? The vast majority of experimental and observational work links
declines in function to declines in functional or species diversity (7,
18, 74). By those criteria very few communities in our dataset are in a
state of concern for loss of functionality (Table \ref{tab:trendTab}),
though we did find evidence that mammal communities may be in greater
danger than bird communities. Still, shifts in metrics are only relevant
if the underlying traits are those most critical for ecosystem function.
We were limited in this analysis to the traits available rather than
those with strong empirical links to function. Similarly, the dimensions
of functional space most important for ecosystem function are still a
topic of on going debate, and at least some known aspects important for
multifunctionality were not measured here (e.g.~dispersion, rarity,
abundance of dominant species, 75).

\hypertarget{potential-methodological-limitations}{%
\subsection{Potential Methodological
Limitations}\label{potential-methodological-limitations}}

Here we approach the question of functional change using the best
available data and biodiversity synthesis approaches. While we found
signatures of change at the study level, some gaps in best practices may
be obscuring a true general trend at the aggregate level. First, the
BioTIME database, while the most comprehensive data source of time
series available, is limited in temporal and geographic scope. Most time
series span only a few years (Fig \ref{fig:taxaMap}) and may not provide
the statistical power necessary to detect trends (76). The database is
also not a representative sample of the world's biodiversity or areas of
greatest threat (6, 12), and the subset of data in this study is
predominantly from North America. We may simply not have data from those
areas experiencing the greatest perturbation (77), particularly
scenarios of conversion to urban, human-dominated landscapes. While
evidence from other work shows even disturbed communities can maintain
functional structure (32, 34), these results should not be interpreted
as evidence of low functional impact in areas of heavy human
disturbance.

Second, despite using the most comprehensive trait databases for these
taxa, we were still limited to species-level means of the traits deemed
important by database creators. The importance of intraspecific
variation is well documented (78, 79), however individual-level traits
are rarely collected alongside monitoring data, especially for the
longest running efforts. Species-level traits may be obscuring more
subtle shifts in the trait space happening within species. Likewise,
available trait data may not capture the traits experiencing the
greatest change. For example, we focus here mostly on ecological traits
(i.e.~foraging strategy, diet, etc) rather than life-history or
reproductive traits.

Third, while we use here the most common metrics for describing
functional diversity, they do not measure some potentially important
aspects of the functional space. Most notably, the summary metrics we
calculated do not capture shifts in the location of the functional space
as a whole. For example, two communities could have very similar metric
values but no overlap in their trait spaces. This is especially relevant
in the context of biodiversity change as a species loss could be
replaced by a species with very different functional attributes, but the
replacement would go undetected if the new species expanded the trait
space by the same degree and had similar abundance. This scenario may be
common in communities tracking changing environmental conditions. While
trait CWM's capture axis shifts, approaches for assessing
multidimensional shifts in functional space are still relatively new
(54, 80, 81) but could shed critical insight into functional composition
changes of this nature.

@@

While we found no over all trends in functional metrics, our results
should not be interpreted as an indication that the ongoing biodiversity
crisis is less severe than previously described, or that there is no
concern for functional change as a result of anthropogenic impact. In
fact, study-level trends indicate quite the opposite, that functional
shifts with unknown implications for ecosystem processes may be going
undetected by common species-based approaches, particularly for short
observation time windows. While the majority of studies in our dataset
did not experience a significant functional loss, a substantial body of
work links functional degradation to species losses as a result of
direct human intervention in the form of land use change and
intensification or habitat fragmentation, indicating that those studies
are simply not representative of the kinds of impacts of greatest policy
concern (31, 82, 83).

>>>>

The meaning of general trends derived from limited data, and their
relevance for conservation, is a topic of on going debate. Global
analyses have been heavily criticized for geographic biases, lack of
data in the most heavily impacted areas, and exclusion of individual
studies' ecological context (10--12). Many of these criticisms reflect
limitations of ecological data on the whole,
\textcolor{blue}{leading to a call for additional data not only to fill geographic and temporal gaps, but to flesh out key characteristics of communities}
(13, 14).

Functional diversity offers a potentially powerful addition to
species-based approaches for detecting and describing community change
by providing a mechanistic link between species' response to
environmental change (\emph{response traits}) and the processes they
perform (\emph{effect traits}) (15--17). By describing the functional
trait space, functional diversity metrics capture the disproportionate
impact of losses or gains of functionally unique species. Functional
diversity metrics
\textcolor{blue}{may therefore illuminate joint responses from functionally similar species or communities undetectible by looking at species identity alone.}

\textcolor{blue}{The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends}
(13). While functional loss is frequently cited as one of the most
pressing concerns of the anthropocene (18--20),
\textcolor{blue}{functional diversity may be maintained even when species are lost from a community}
(21, 22). Forecasts of functional loss range from negligible (23) to
dire (24, 25). And while some observed trends show significant
functional loss (26) others document no loss even in some of the most
heavily impacted communities (27--29). On paleoecological time scales,
functional composition shows mixed responses to environmental change and
extinction events (30, 31), with significant impacts of species
extinctions on functional diversity in some taxa and not others (32).
For some time periods, functional structure appears to be maintained for
substantial portions of geological time (33).
\textcolor{blue}{Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds [@barnagaud2017; @jarzyna2016] and ray-finned fishes, sharks, and rays}
(34).

Here we
\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and computational advances to}
perform the first multi-taxa, multi-realm assessment of functional
diversity change through time. We focus on mammal and bird species as a
significant subset of the world's biodiversity heavily impacted by
anthropogenic change. While examining trends in plants, invertebrates,
and other vertebrate species is of equal interest, trait data for those
taxa raise additional challenges such as limited and biased species
coverage (35), a lack of accepted species-level means, and differences
in the types of traits collected. To ensure comparability across taxa in
trait type and data quality we therefore focus on mammals and birds.
Traits were intentionally selected to be representative of a species'
Eltonian niche, thereby summarizing the functional role they play in the
community (36). An initial assessment of amphibian trends is included in
the supplement, but excluded from general trend assessment here due to
limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series
to determine whether or not
\textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities. Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination. We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm), and trends for individual studies.}

@@

We obtained mammal and bird time series from the BioTIME database, a
global repository of high-quality assemblage time series. All studies
included in the database follow consistent sampling protocols and
represent full assemblages rather than populations of single species
(31). All time series include abundance of observed species. Following
best practices for the database (37), studies with multiple sample
locations were split into individual time series following a
standardized spatial scale. Scale was set by a global grid with cell
size determined based on the sample extent of studies with only a single
location (see 31 for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations. All samples from a study within a single
cell were considered to be a single time series, and species abundances
were combined for all samples.

We used trait data from the Elton Trait Database, which consists of
species-level means for traits that represent species' multifaceted role
in the community (36). Traits include: body mass, diet, active diel
period, nocturnality, forest foraging strata, pelagic use. Multiple
traits (i.e.~diet, foraging strata, activity seasonality, active diel
period) were broken down into percentage or binary use for each level.

In order to ensure taxonomic consistency across datasets, BioTIME
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (retrieved
09-15-2020 from the on-line database,
\url{https://doi.org/10.5066/F7KH0KBK}), obtained through the
\texttt{taxadb} R package (38, 39). If more than one species in the
assemblage data resolved to the same identifier, observations were
considered the same species. For trait data, traits for all species of
the same identifier were averaged. Only studies for which at least 75\%
of species had trait data were included. In order to have a sufficient
number of species to calculate functional diversity metrics, years with
fewer than 5 species observed were also excluded. Sensitivity analyses
were conducted for the trait coverage threshold and the duration of
included time series.

@@

Functional diversity metrics were calculated using the \emph{dbFD}
function from the \emph{FD} R package (40). Here we report functional
richness (\emph{FRic}), functional evenness (\emph{FEve}), and
functional divergence (\emph{FDiv}) which together describe three
complementary characteristics of the functional space (41, 42).
\emph{FRic} assesses the volume of the trait space occupied by species
in the community, with higher values indicating communities with species
of more extreme trait values. \emph{FEve} describes how species are
distributed across the trait space and how abundance is distributed
across species. Higher values of \emph{FEve} indicate more even spacing
of species in the trait space and individuals across species.
\emph{FDiv} measures the degree to which species and their abundances
maximize differences in the functional space. Higher values of
\emph{FDiv} therefore correspond to communities where many highly
abundant species are on the edges of the trait space. We also calculated
the community-weighted mean (\emph{CWM}) of included traits to examine
shifts in the distribution of each trait.
\textcolor{blue}{Hereafter we refer to results for functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).}

All available trait data for each study were included in functional
diversity calculations with the exception of traits that were the same
value for all observed species in the study. For variables with multiple
levels each level was included as a separate trait axis. Continuous
traits were z-score scaled to give each trait equal weight in the trait
space (43, 44).
\textcolor{blue}{Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA. We limited the number of included PCoA axes to}
the maximum number of traits that fulfills the criteria \(s >= 2^t\),
where \(s\) is the number of species and \emph{t} is the number of
traits. This restriction allows for enough axes to capture the trait
space while maintaining computational feasibility (45). Metrics
incorporated weighting based on species abundance.

@@

To assess functional change independent of species richness we
calculated the standardized effect size (SES) for each of the three
functional diversity metrics (\emph{FRic}, \emph{FEve}, \emph{FDiv})
from null estimates (46). Null model corrections allow us to assess the
degree to which the observed functional diversity metric deviates from
the value expected by chance in a randomly assembled community. Null
estimates were calculated for each rarefied sample by randomly sampling
species from the species pool for each year and randomly assigning
observed abundances to species. `
\textcolor{blue}{Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability}.
This process was repeated 500 times to get an estimate and standard
deviation of the null expectation for the metric for each rarefaction
sample for that time series. We used these values to calculate SES using
the following formula:
\(SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}\). We then
calculated the median SES estimate for each metric from all the
rarefaction samples for a time series. SES estimates can be interpreted
as how much of the functional characteristic (richness, evenness,
divergence) was observed beyond what was expected by chance for a
community of that species richness.

@@

We estimated general trends
\textcolor{blue}{across bird and mammals communities} for each diversity
metric using a linear mixed effects model with a random slope and
intercept for each study and each time series nested within the study.
We fit 18 individual \emph{CWM} models. All time series with data for a
given trait were included in the corresponding \emph{CWM} model. We
estimated study level trends using individual linear models. For studies
with more than one times series we fit a random slope and intercept for
time series. Some study-level models could not be fit for five studies
for at least one metric due to data limitations, but those studies were
still included in the general models. They represented 12 of 1350
study-level models fit for each metric. For further details see the
supplement. Where appropriate, response variables were \(log\) or
\(log(x+1)\) transformed to better fit model assumptions of residual
normality.

To test for trends within and between different levels of taxa, biome,
and realm we fit separate models with each of those factors added as a
predictor to the original model structure. We estimated within-level
slopes and calculated between-level contrasts using the \emph{emmeans}
package (v1.8.2, 47). We assessed the impact of time series duration and
start year on study-level trends using linear models with duration and
start year as predictors. All models in our analysis were fit using the
\emph{lme4} (v1.1-30) package in R (v4.2.3) and p-values were calculated
by Satterthwaite's degrees of freedom method using the \emph{lmerTest}
(v3.1-3) package with a significance level of \(\alpha = 0.05\)
(48--50).

@@

We found no significant overall trend in species richness or functional
diversity metrics (observed or \textcolor{blue}{corrected}) (Fig
\ref{fig:timeseriesPlot}). We did find a significant overall decrease in
Jaccard similarity, indicating accumulating changes in species
composition. Non-significant overall trends indicate that although some
studies experience increasing or decreasing trends, the average trend
across studies was plausibly 0 (Table \ref{tab:resultsTab}).
\textcolor{blue}{Within-group} trends for different taxa, biomes, or
realms were also non-significant for richness and functional diversity
metrics, with the exception of a significantly increasing trend for
functional evenness of global studies (characterized by having samples
on multiple continents), and a significantly decreasing functional
richness slope for temperate/tropical studies and mammal studies.
However, trends were not exhibited for the \textcolor{blue}{corrected}
metric indicating that differences
\textcolor{blue}{in functional diversity metrics} were largely due to
changes in species richness. Further, with only two global studies, the
trend should not be considered truly general. The general trends for
\emph{CWM} models were similarly not significant, with a significant
positive trend for only percentage of fish in diet composition.

@@

At the study level, 15 of 50 studies exhibited a significant trend in
species richness and 11 exhibited significant turnover. For observed
functional metrics, 25 of 50 studies exhibited a trend in a least one
metric, and 15 of 50 studies exhibited a significant trend for at least
one \textcolor{blue}{corrected} metric (Table \ref{tab:trendTab}). In
general, there were more significant trends for uncorrected metrics,
with some disappearing after correction, indicating that those trends
were likely due to changes in the number of species. Hypothesis testing
for study-level trends is likely affected by multiple testing issues and
some trends identified as significant are therefore potentially
spurious. Rather than interpreting changes in specific studies, we
present these results as a general picture of the kinds of trends
experienced by communities.

Study-level slopes for functional diversity metrics were significantly
related to start year of the time series for Jaccard similarity and
functional evenness, both of which had significantly more negative
slopes with more recent start year. No functional diversity metrics were
significantly related to the start year of the time series.

@@

Our study represents the largest broad-scale multi-taxa assessment of
functional change through time to date, giving a first look at aggregate
and local trends in functional diversity in mammal and bird communities.
\textcolor{blue}{Our results show that the addition of functional traits illuminates a few consistent functional trends across communities, but largely complicates rather that clarifies the story of biodiversity change. While the characteristics of species clearly matter, instead of unifying the nature of communities' change, they more often distinguish them. In general, we found a few areas of consensus for models where communities were aggregated (general trends, trends by taxa, biome, and realm), and vast heterogeneity for study-level models.}

\hypertarget{evidence-of-consensus}{%
\subsubsection{Evidence of consensus}\label{evidence-of-consensus}}

We did not detect an overall trend in any functional diversity metrics.
As with previous species-based syntheses, we also found no overall trend
in species richness accompanied by increasing turnover through time
(31), indicating that non-significant trends in functional metrics may
be consistent with similar well-documented species derived trends. We
found no trend in functional change for almost all realms, biomes, and
taxonomic groups.
\textcolor{blue}{The one exception was a consensus trend indicating}
functional richness loss for mammal studies. These results are
consistent with multiple studies linking anthropogenic drivers to loss
of functional diversity in mammal communities (51--53).

Despite a lack of general trends in functional diversity metrics, we did
find evidence of compositional shifts in the trait space with multiple
\texttt{significant} trends in \emph{CWM}'s for studies coming from the
same taxa, realms, or climates,
\texttt{indicating\ areas\ of\ consensus\ change\ across\ communities}.
Some of those trends are consistent with existing work in those
contexts. For example, the decrease in insect consumption for tropical
communities reflects well documented declines in tropical insectivorous
birds (54). Reduction in utilization of understory foraging for
terrestrial communities could be the result of disturbance like human
recreation or increased predation from introduced species. Other shifts
in diet for both Birds and Mammals, including increasing fruit, seed,
and vertebrate consumption and decreasing seed consumption in some
climates point to important areas for further exploration.

\hypertarget{study-heterogeneity}{%
\subsubsection{Study heterogeneity}\label{study-heterogeneity}}

The strongest picture the study-level results give is that trait level
data further complicates the story of biodiversity change. The lack of
general trend in functional diversity metrics belies a huge range of
positive and negative trends at the study-level that happen to average
out to zero. In order to simplify discussion, we will talk about the
implications for communities grouped by their concurrent change in
functional diversity metrics, species richness and turnover split into
positive, negative or no trend. While there are over 150 possible
combinations of change direction (or no change) in the 5 metrics, we
discuss here the six scenarios that occurred in more than one study: no
change, species loss or gain only, loss of functional evenness, richness
loss with species turnover, and increase in species and functional
richness accompanied by significant turnover. We focus on these
scenarios to combat potential spurious results due to multiple testing,
as it is unlikely all observations of the same scenario are spurious.
The number of studies in each scenarios is broken down in table x to
illustrate the relative number between groups rather than absolute
number.

The majority group of studies exhibited no trend in any species or
functional diversity metric.

@@

About 15 percent of studies, which were also exclusively the longest
running studies, had a trend in turnover, richness, and functional
change. The studies in this scenario fell into two groups: a positive
trend in richness with a positive trend in functional richness
(corrected and uncorrected), and a negative trend in species richness
with a negative trend in uncorrected functional richness and no trend in
corrected richness but a trend in other corrected functional metrics.
\textcolor{blue}{This would indicate that when species are gained in a community, they disproportionately add to the functional richness. When species are lost however, they are not more functionally unique than expected by chance though they may produce shifts in other aspects of functional space.}

Our results illustrate the range of relationships between species-based
and functional metrics observed in real communities. Critically,
particularly for the short observation time windows that characterize
the majority of ecological data, functional shifts may be underlying
seemingly static species-based community metrics. Species-based
approaches may therefore be missing areas where significant change in
the functional space could be impacting ecosystem function and
resilience. In scenarios where there are significant species gains or
losses over time, we found evidence that species losses may be
relatively functionally indistinct, while species gains contribute
disproportionately to the richness of a functional space. This finding
is consistent with evidence that at least in bird communities, common
functionally general species are being lost even as rare species are
increasing (55--57).

Our results are consistent with ecological expectations that community
dissimilarity would be greater the larger the observation time window
purely due to background processes, which is likely further accelerated
by global change drivers (13). While we do not address how observed
rates differ from background expectation here, we can make some
assertions about the implication of turnover for functional change.
Where turnover is occurring, it is almost always accompanied by a
\textcolor{blue}{trend in one of the functional diversity metrics, indicating}
directional change in the functional space. Species are not therefore
being replaced by exact functional counter parts. The fact that we
observed both increases and decreases in corrected functional richness
accompanying turnover is a reminder that
\textcolor{blue}{new species are not always replacing more functionally unique species, but are often more functionally unique themselves.}

@@

Still, some results stand in contrast to predictions for trait shifts
under global change. For example, mean body size is predicted to
decrease as a result of climate change impacts and megafaunal loss (58),
a phenomenon which has already been well documented empirically and
experimentally in multiple taxa (59--62). While the species-level trait
means used here are not appropriate for assessing intraspecific body
size shifts, we would expect to see shifts in community-wide means due
to local losses of large-bodied species. Instead, we found no evidence
of a trend in \emph{CWM} body size, with the exception of a single study
of marine mammals in the Bahamas which showed a significant increase.
Still, many of the studies in our dataset draw from areas that may have
experienced significant loss of large-bodied species before the
observation window, with contemporary loss rates slowing (63). Trends
could be significantly different for the same time periods in regions of
sub-Saharan Africa, for example, which has poor representation in our
dataset but where megafauna exist on the landscape and are increasingly
threatened (64).

What does local maintenance of functional structure mean for ecosystem
function? The vast majority of experimental and observational work links
declines in function to declines in functional or species diversity (7,
65, 66). By those criteria very few communities in our dataset are in a
state of concern for loss of functionality (Table \ref{tab:trendTab}),
though we did find evidence that mammal communities may be in greater
danger than bird communities. Still, shifts in metrics are only relevant
if the underlying traits are those most critical for ecosystem function.
We were limited in this analysis to the traits available rather than
those with strong empirical links to function. Similarly, the dimensions
of functional space most important for ecosystem function are still a
topic of on going debate, and at least some known aspects important for
multifunctionality were not measured here (e.g.~dispersion, rarity,
abundance of dominant species, 67).

@@

While we found no over all trends in functional metrics, our results
should not be interpreted as an indication that the ongoing biodiversity
crisis is less severe than previously described, or that there is no
concern for functional change as a result of anthropogenic impact. In
fact, study-level trends indicate quite the opposite, that functional
shifts with unknown implications for ecosystem processes may be going
undetected by common species-based approaches, particularly for short
observation time windows. While the majority of studies in our dataset
did not experience a significant functional loss, a substantial body of
work links functional degradation to species losses as a result of
direct human intervention in the form of land use change and
intensification or habitat fragmentation, indicating that those studies
are simply not representative of the kinds of impacts of greatest policy
concern (26, 68, 69).
